A thermodynamic view of dusty protoplanetary disks

### 2017-05-02 model.tex
lag cartoon update, minor text edits
misc edits from sect 4 to sect 8.3

The total gravitational potential is $\Phi_\mathrm{tot}=\Phi + \psi$,
where $\Phi$ originates from a star of mass $M_*$, 
\begin{align}\label{thin_disk_potential}
  \Phi(r,z) =-\frac{GM_*}{\sqrt{r^2 + z^2}}\simeq
  -\frac{GM_*}{r}\left(1 - \frac{z^2}{2r^2}\right), 
\end{align}
and $\psi$ is the disk potential satisfying
\begin{align}
  \nabla^2\psi = 4 \pi G \rho.
\end{align}
Here, $G$ is the gravitational constant, and $(r,\phi, z)$ are 
cylindrical co-ordinates centered on the star.  
The second equality in Eq. \ref{thin_disk_potential} is the 
thin-disk approximation for the stellar potential, appropriate for
$|z|\ll r$. We shall adopt this approximate potential in order to
obtain explicit expressions for disk equilibria (\S\ref{eqm}).

The gravitational potential of the mixture 
$\psi$ is included here for completeness, but we will mostly neglect
it ($\psi \equiv 0$) unless explicitly stated otherwise.

@@

\subsection{Effective energy equation}
Although we have eliminated the need of an evolutionary energy
equation (Eq. \ref{realenergy}) by choosing a locally isothermal
equation of state (Eq. \ref{eos}), we show that the mixture 
nevertheless obeys an effective evolutionary energy equation. This is
because advection of the dust-fraction, described by
Eq. \ref{dusteq}, can be transformed into an energy-like 
equation. We 
%We show that for a prescribed temperature distribution, the mixture
%obeys an evolutionary energy equation, due to the advection of the
%dust-fraction. 
eliminate $\tepsilon$ in favor of $P$, 
\begin{align*}
  \tepsilon = 1 - \frac{P}{c_s^2(r,z)\rho}, 
\end{align*}
then Eq. \ref{dusteq} becomes
\begin{align}
%\frac{DP}{Dt} 
\frac{\p P}{\p t} + \bm{v}\cdot\nabla P  
&= - P \nabla\cdot\bm{v} + P\bm{v}\cdot\nabla\ln{c_s^2}
                + \mathcal{C},  \label{eff_energy} \\
\mathcal{C}&\equiv c_s^2 \nabla\cdot\left[\tstop\left(1 -
  \frac{P}{c_s^2\rho}\right)\nabla 
  P\right].
\end{align}
Comparing with Eq. \ref{realenergy}, we see that 
Eq. \ref{eff_energy} is the energy equation for an ideal gas of
adiabatic index $\gamma=1$, but now with the imposed temperature
gradient $\nabla c_s^2$ and dust-gas drag $\mathcal{C}$ playing the
role of radiative cooling ($\Lambda$ in 
Eq. \ref{realenergy}).

@@

For a given distribution of the dust-fraction $\tepsilon$ (or
dust-to-gas ratio $\epsilon$), the 
mass and momentum Eqs. \ref{masseq}---\ref{momeq} admit     
axisymmetric steady states with $\rho(r,z)$ and 
$\bm{v}=r\Omega(r,z)\hat{\bm{\phi}}$ where $\Omega = v_\phi/r$, which satisfy 
\begin{align}
  r\Omega^2 &= \frac{\p \Phi}{\p r} + \frac{1}{\rho}\frac{\p P}{\p
    r},\label{steady_momr}\\
  0 & = \frac{\p\Phi}{\p z} + \frac{1}{\rho}\frac{\p P}{\p z},\label{steady_momz}
%  0 & = \nabla\cdot\left(\tepsilon\tstop\nabla P\right) \label{steady_dust}
\end{align}
with $P=P(\tepsilon,\rho)$ given by the equation of state
(Eq. \ref{eos}). An explicit solution is presented in
\S\ref{steady_state}.  
However, disk structures obtained this way generally do not satisfy 
the steady state effective energy Eq. \ref{eff_energy} because
$\mathcal{C}\neq0$ for general solutions to
Eq. \ref{steady_momr}---\ref{steady_momz}.

In practice, dust particles settle to the disk midplane on a timescale 
$t_\mathrm{settle}\sim 1/\OmK^2\tstop$, although for tightly-coupled dust
particles this is much longer than that to establish vertical
hydrostatic equilibrium in the gas component ($\sim 1/\OmK$).  
True steady states may be calculated by invoking an underlying gas
turbulence, which leads to dust diffusion counter-acting vertical
settling \citep{takeuchi02, youdin07}. This complication is beyond the
scope of this paper.

In this work, consider equilibria with $\mathcal{C}\equiv 0$, namely   
\begin{inparaenum}[1)] 
\item 
  unstratified disks with finite dust-gas drag; %(\S\ref{si});  
\item 
  perfectly-coupled dust with $\tstop=0$. %(\S\ref{results}). 
\end{inparaenum} 
Vertical settling is absent in both cases. Alternatively, we are 
assuming that $\tstop$  is sufficiently small such 
that settling occurs on much longer timescales than the instabilities of interest.

@@

\subsection{Vertical buoyancy}\label{vbuoyancy}
Having identified the entropy of an isothermal dusty gas, we find the
vertical buoyancy frequency $N_z$ of the mixture is given by 
\begin{align}
  N_z^2 \equiv - \frac{1}{\rho}\frac{\p P}{\p z}\frac{\p \seff}{\p z} &=
  \frac{c_s^2(r)}{\left(1+\epsilon\right)^2}\frac{\p\ln\rhog}{\p 
  z}\frac{\p\epsilon}{\p z} \\ &
                                  =
  \frac{\epsilon}{\left(1+\epsilon\right)^2}\left(\frac{z}{\Htilde}\right)^2\OmK^2\notag,  
\end{align}
where the second equality assumes $c_s=c_s(r)$ and the final equality
applies to the equilibria in \S\ref{eqm}. Thus,  
\begin{align*}
N_z\lesssim
O\left(\sqrt{\epsilon_0}\OmK\right). 
\end{align*}
However, for well-mixed dust layers such that $H_\epsilon \gg \Hg$, 
$\mathrm{max}\left(N_z\right)$ may occur outside a finite vertical domain.

@@

In this section we discuss the basic properties of the 
dusty-gas mixture using Eqs.\ref{masseq}--\ref{momeq} with 
Eq. \ref{eff_energy} in place of the dust continuity equation. Our
discussion is based on the corresponding axisymmetric linearized
equations. A variable $X$ is
subject to Eulerian perturbations of the form 
\begin{align}
 \real\left[ \delta X(r,z)\exp{\left(-\ii\sigma t\right)}\right], 
\end{align}
where $\sigma$ is the complex mode frequency. For later convenience we
write 
\begin{align}
  \sigma = \ii s - \omega,
\end{align}
where $s$ and $\omega$ are the growth rate and real frequencies,
respectively. Thus perturbations have time dependence $e^{st +
  \ii\omega t}$.

@@

In this case, the last two integrals in Eq. \ref{int_rel} vanish. 
The condition for axisymmtric \emph{stability}, $\sigma^2>0$, 
is met if the integrand cofficients satisfy $A+D>0$ and $AD-B^2>0$
\citep[][Section 11.6]{ogilvie16}, which becomes 
\begin{align}
  \kappa^2 + c_s^2 \nabla\ln{\rhog}\cdot\nabla\tepsilon &> 0,\label{dusty_solberg1}  \\
  -c_s^2\frac{\p\ln{\rhog}}{\p z}\left(-\kappa^2\frac{\p\tepsilon}{\p
    z} + r\frac{\p\Omega^2}{\p z}\frac{\p \tepsilon}{\p r} \right) & > 0, \label{dusty_solberg2}
\end{align} 
%\begin{align}
%  \kappa^2 - \frac{1}{\rho}\nabla P \cdot \nabla s &> 0, \label{dusty_solberg1}    \\
%  -\frac{1}{\rho}\frac{\p P}{\p z} \left(\kappa^2 \frac{\p s}{\p z} -
%  r\frac{\p\Omega^2}{\p z}\frac{\p s}{\p r}\right) &>0, \label{dusty_solberg2}
%\end{align} 
where %$s = \ln{P/\rho}$, 
$\kappa^2 \equiv r^{-3}\p_r\left(r^4\Omega^2\right)>0$ is the 
square of the epicylic frequency. %We assume $\kappa^2>0$.  
Eq. \ref{dusty_solberg1}---\ref{dusty_solberg2} are in fact the 
the Solberg-Hoiland criteria for axisymmetric stability of adiabatic
gas if one inserts $C_p\ln{\left(1 -
    \tepsilon\right)}$ for the entropy in the standard
expression for said criteria \citep{tassoul78}.

We expect Eq. \ref{dusty_solberg1} to be satisfied in typical
protoplanetary disks where the dust-to-gas ratio increases in the same
direction as the local pressure gradient, which is equivalent to the
gas density gradient for isothermal gas. Thus to enable instability
here we need to violate Eq. \ref{dusty_solberg2}. We explore this
below.

\subsubsection{Vertically well-mixed dust}\label{vert_mixed_theory}
Here we consider the special case where the dust-to-gas ratio is
uniform in $z$. Then the LHS of Eq. \ref{dusty_solberg2} becomes
\begin{align}
-\frac{c_s^4}{(1+\tepsilon)^4}\left(\frac{\p\ln{\rhog}}{\p
    z}\right)^2\left(\frac{\p\epsilon}{\p r}\right)^2<0.
\end{align}
Thus the disk is \emph{unstable}, since there is no effective vertical 
buoyancy to stabilize vertical motions. This situation is
identical to the pure gas, adiabtic simulations of \cite{nelson13}
where the gas entropy is vertically uniform but has a radial
gradient. The authors indeed find instability. We give a numerical
example in the dusty context in \S\ref{vert_mixed}.

\subsubsection{Order-of-magnitude estimates} 
In order to apply the stability criteria
Eq. \ref{dusty_solberg1}---\ref{dusty_solberg2} to typical
protoplanetary disks we conside the equilibria described in
\S\ref{eqm}.

We assume the disk is approximately Keplerian so that $\kappa\simeq\OmK$, and that the vertical gas distribution is
Gaussian. In terms of the dust-to-gas
ratio $\epsilon$, Eq. \ref{dusty_solberg1}---\ref{dusty_solberg2}
become   
\begin{align}
  &1 + \frac{\epsilon}{\left(1+
    \epsilon\right)^2}\left(h_\mathrm{g}^2\frac{\p\ln{\rhog}}{\p\ln{r}}
  \frac{\p\ln{\epsilon}}{\p\ln{r}} + 
  \frac{z^2}{\Htilde^2}\right)>0 \label{stability_est1},   \\ 
&1 - \frac{\epsilon
  h_\mathrm{g}^2}{\left(1+\epsilon\right)^2}
  \frac{\p\ln{\epsilon}}{\p\ln{r}}\left(-\frac{\p\ln{\rhog}}{\p\ln{r}}+\frac{\Htilde^2}{H_\mathrm{g}^2}\frac{\p\ln{\epsilon}}{\p\ln{r}}\right)
  > 0 \label{stability_est2},
\end{align}
\emph{for stability}.

Eq. \ref{stability_est1}---\ref{stability_est2} are easily satisfied
in thin accretion disks where radial gradients are $O(1/r)$, 
$h_\mathrm{g}\ll 1$ and $\Htilde/H_\mathrm{g}$ not too
large.  Then the
magnitude of the second term on the left-hand-side  of either
inequality is much less than unity. Thus radially smooth, strictly
isothermal disks perfectly coupled with dust are stable against
axisymmetric perturbations. This result holds for arbitrarily thin
dust layers.

%Such disks are stable against
%axisymmetric perturbations.

\subsubsection{Instability at dust edges}
Notice the left-hand-side of Eq. \ref{stability_est2} is a quadratic in
$\p_r\epsilon$. It is possible to
violate this inequality for sufficiently large (in magnitude) radial
gradients  in the dust-to-gas ratio, 
\begin{align}
  \frac{\p\ln{\epsilon}}{\p\ln{r}} > \chi_+ \quad \mathrm{or} \quad 
  \frac{\p\ln{\epsilon}}{\p\ln{r}} < \chi_-,
\end{align}
\emph{for instability}, where
\begin{align}\label{spm}
\chi_\pm = \frac{1}{2}\frac{H_\mathrm{g}^2}{\Htilde^2} 
  \left[
  \frac{\p\ln{\rhog}}{\p\ln{r}} \pm 
  \sqrt{
  \left(\frac{\p\ln{\rhog}}{\p\ln{r}}\right)^2 + 
  4 \frac{\Htilde^2}{H_\mathrm{g}^2}
  \frac{\left(1+\epsilon\right)^2}{\epsilon h_\mathrm{g}^2}
  }
  \,\right]. 
\end{align} 
In typical accretion disks where $\p_r\rhog<0$, it is easier to
achieve instability at a given radius for increasing dust-to-gas
ratios in outwards ($\p_r\epsilon > 0$), and vice versa.

We can neglect density gradients in Eq. \ref{spm} 
if $r\p_r\ln{\rhog}\sim O(1)$ and $H_\epsilon/\Hgas\gg
\sqrt{\epsilon}\hgas$. For example, if $\epsilon\simeq 0.01$ and
$\hgas\simeq 0.05$, then we require $H_\epsilon/\Hg\gg
5\times10^{-3}$. (So the dust layer thickness $\Hd\gg 5
\times10^{-3}\Hgas$ as well.) Then 
instability requires 
\begin{align}\label{dust_edge}
\left|\frac{\p\ln{\epsilon}}{\p r}\right| \gtrsim
  \frac{1}{\Htilde}\frac{\left(1+\epsilon\right)}{\sqrt{\epsilon}}.
%\simeq
 % \frac{1}{\sqrt{\epsilon}H_\epsilon}, 
\end{align}
%where the second equality applies to $\epsilon\ll 1$. 
That is, if the radial lengthscale of the dust-to-gas ratio,
$L_\epsilon\ll O(\Htilde)$, i.e. its vertical lengthscale,  
then the system is potentially unstable.

Taking $\epsilon\sim 0.01$,  we find that 
for thin dust layers with $H_\epsilon \simeq \Hd\ll \Hg$,  instability
requires $L_\epsilon\ll O(0.1\Hg)$, i.e. the dust-to-gas ratio must
vary on an extremely short lengthscale. 
% %an extremely short  
% %radial length-scale in the dust-to-gas ratio. 
% This is unlikely to
% occur. On the other hand, if dust is well-mixed, say $\Hd\simeq
% 0.99\Hg$, then instability only requires 
% $L_\epsilon\lesssim O(10\Hg)$, which is more realistic.

%This is responsible for 
%the usual vertical shear instability in locally isothermal disks.
%However, we may obtain an integral
%relation for linear, axisymmetric waves with frequency $\sigma$ (see
%Appendix \ref{var_prin}, where we also account for finite $\tstop$) in
%the form:   
%If $c_s$ is constant, then $\sigma^2$ is real, and 
%the first integral leads to the previous Solberg-Hoiland criteria for 
%axisymmetric stability. However, when $c_s$ is non-uniform,

\subsection{Thermodynamics of dust-drag instabilities}\label{dust_work}
Now consider constant $c_s$ but $\tstop\neq0$. 
Eq. \ref{int_rel} indicates that $\sigma^2$ is generally complex, so
we may have growing oscillations. %Writing $\sigma =
%-\omega + \ii s$ where $\omega$ and $s$ are the real frequency and
%growth rates, respectively, 
The imaginary part of Eq. \ref{int_rel}
gives  
\begin{align}
%  s = \frac{\imag\int \left(\nabla\cdot\dd\bm{v}^*\right)\dd\mathcal{C}dV}{2\omega\int\rho\left(|\dd
%    v_r|^2 + |\dd v_z|^2\right)dV}, \label{thermal_instability}
  s = \frac{\imag\int \left(\nabla\cdot\dd\bm{v}^*\right)\dd\mathcal{C}dV}{2\omega\mathcal{I}^2}, \label{thermal_instability}
\end{align} 
assuming $\omega\neq0$. This represents overstability due to dust-gas friction. 
Since drag ($\delta \mathcal{C}$) appears as a source term in our
effective energy equation, the quantity 
$\imag\left[\left(\nabla\cdot\dd\bm{v}^*\right)\dd\mathcal{C}\right]$
may be thought of as a correlation between compression/expansion and  
heating/cooling.

@@

\subsubsection{Physical requirement of dust-gas drag overstabilities}

@@

A parcel of the isothermal dusty-gas mixture does 
postive work if  
\begin{align*}
-\sgn\left(\omega\right)\imag\left(\Delta\rhog\Delta\rhod^*\right)>0,
\end{align*}
meaning that gas follows dust. Instabilities are thus not possible if the
gas does not respond to dust-drag (i.e. no back-reaction). The
pressure-density lag shown in Fig. \ref{pdv_cartoon} is achieved if, 
after a parcel reaches maximum total density, its 
gas content is still increasing ($A$ to $B$). This in turn implies a sufficiently
large particle flux \emph{out} of the parcel.

@@

This thermodynamic interpretation does not explain 
\emph{why} drag forces causes gas pressure to lag behind the dust
density, but shows 
that this \emph{must} be the case for any growing modes associated the
dust-gas drag. To rigorously understand how dust-drag causes this 
lag requires an explicit solution to the linearized equations with 
detailed treatment of the function $\mathcal{C}$. However, given the complexity of
$\mathcal{C}$ (see Appendix \ref{lin_dust}), we might generally expect
 dusty disks to  support a range of stable and unstable modes, with the
latter being associated with pressure-density lag.

Indeed, \cite{jacquet11} explains the essence of the streaming 
instability in dusty protoplanetary disks as dust accumulation
(and hence compression) at a pressure bump, which then drags the 
gas towards it (and hence heating) to strengthen the pressure
bump. That is, heating occurs upon compression, as in stellar  
pulsational instabilities \citep{cox67}. In \S\ref{si} we check that the
streaming instability fits into this thermodynamic interpretation in
the strong drag limit.

>>>>

The total gravitational potential is $\Phi_\mathrm{tot}=\Phi +
\psi$. Here $\Phi$ originates from a star of mass $M_*$, 
\begin{align}\label{thin_disk_potential}
  \Phi(r,z) =-\frac{GM_*}{\sqrt{r^2 + z^2}}\simeq
  -\frac{GM_*}{r}\left(1 - \frac{z^2}{2r^2}\right),  
\end{align}
where $(r,\phi, z)$ are cylindrical co-ordinates centered on the star.   
The second equality in Eq. \ref{thin_disk_potential} is the 
thin-disk approximation for the stellar potential, appropriate for
$|z|\ll r$. We shall adopt this approximate potential in order to
obtain explicit expressions for disk equilibria (\S\ref{eqm}).

The disk's own gravitational potential $\psi$ is included for
completeness, but we will mostly neglect it unless stated
otherwise. In such cases $\psi$   
satisfies  
\begin{align}
  \nabla^2\psi = 4 \pi G \rho,
\end{align}
where $G$ is the gravitational constant.

@@

\subsection{Effective energy equation}\label{energy_analogy}
Although we have eliminated the need of an evolutionary energy
equation (Eq. \ref{realenergy}) by choosing a locally isothermal
equation of state (Eq. \ref{eos}), we show that the mixture 
nevertheless obeys an effective evolutionary energy equation. This is
because advection of the dust-fraction, described by
Eq. \ref{dusteq}, can be transformed into an energy-like 
equation. We 
%We show that for a prescribed temperature distribution, the mixture
%obeys an evolutionary energy equation, due to the advection of the
%dust-fraction. 
eliminate $\tepsilon$ in favor of $P$, 
\begin{align*}
  \tepsilon = 1 - \frac{P}{c_s^2(r,z)\rho}, 
\end{align*}
then Eq. \ref{dusteq} becomes
\begin{align}
%\frac{DP}{Dt} 
\frac{\p P}{\p t} + \bm{v}\cdot\nabla P  
&= - P \nabla\cdot\bm{v} + P\bm{v}\cdot\nabla\ln{c_s^2}
                + \mathcal{C},  \label{eff_energy} \\
\mathcal{C}&\equiv c_s^2 \nabla\cdot\left[\tstop\left(1 -
  \frac{P}{c_s^2\rho}\right)\nabla 
  P\right].
\end{align}
Comparing with Eq. \ref{realenergy}, we see that 
Eq. \ref{eff_energy} is the energy equation for an ideal gas of
adiabatic index $\gamma=1$, but now with the imposed temperature
gradient $\nabla c_s^2$ and dust-gas drag $\mathcal{C}$ playing the
role of radiative cooling ($\Lambda$ in 
Eq. \ref{realenergy}).

@@

For a given distribution of the dust-fraction $\tepsilon$ (or
dust-to-gas ratio $\epsilon$), the 
mass and momentum Eqs. \ref{masseq}---\ref{momeq} admit     
axisymmetric steady states with $\rho(r,z)$ and 
$\bm{v}=r\Omega(r,z)\hat{\bm{\phi}}$ where $\Omega = v_\phi/r$, which satisfy 
\begin{align}
  r\Omega^2 &= \frac{\p \Phi}{\p r} + \frac{1}{\rho}\frac{\p P}{\p
    r},\label{steady_momr}\\
  0 & = \frac{\p\Phi}{\p z} + \frac{1}{\rho}\frac{\p P}{\p z},\label{steady_momz}
%  0 & = \nabla\cdot\left(\tepsilon\tstop\nabla P\right) \label{steady_dust}
\end{align}
with $P=P(\tepsilon,\rho)$ given by the equation of state
(Eq. \ref{eos}). An explicit solution is presented in
\S\ref{steady_state}.  
However, disk structures obtained this way generally do not satisfy 
the steady state effective energy Eq. \ref{eff_energy} because typically 
$\mathcal{C}\neq0$.  
%for general solutions to 
%Eq. \ref{steady_momr}---\ref{steady_momz}.

% Dust particles settle to the disk midplane on a timescale 
% $t_\mathrm{settle}\sim 1/\OmK^2\tstop$, although for tightly-coupled dust
% particles this is much longer than that to establish vertical
% hydrostatic equilibrium in the gas component ($\sim 1/\OmK$).  
% True steady states may be calculated by invoking an underlying gas
% turbulence, which leads to dust diffusion counter-acting vertical
% settling \citep{takeuchi02, youdin07}. This complication is beyond the
% scope of this paper.

In order to have a well-defined stability problem, we formally
take equilibria with $\mathcal{C}\equiv 0$, namely      
\begin{inparaenum}[1)] 
\item 
  unstratified or 2D disks with finite dust-gas drag; %(\S\ref{si});   
\item 
  perfectly-coupled dust with $\tstop=0$. %(\S\ref{results}). 
\end{inparaenum} 
Alternatively, we are assuming that $\tstop$  is sufficiently small
such that the evolution of the background disk (e.g. dust-settling) 
occur on much longer timescales than the 
instabilities of interest.

@@

\subsection{Dusty vertical buoyancy}\label{vbuoyancy}
Having identified the entropy of an isothermal dusty gas, we find the
vertical buoyancy frequency $N_z$ of the mixture is given by 
\begin{align}
  N_z^2 \equiv - \frac{1}{\rho}\frac{\p P}{\p z}\frac{\p \seff}{\p z} &=
  \frac{c_s^2(r)}{\left(1+\epsilon\right)^2}\frac{\p\ln\rhog}{\p 
  z}\frac{\p\epsilon}{\p z} \\ &
                                  =
  \frac{\epsilon}{\left(1+\epsilon\right)^2}\left(\frac{z}{\Htilde}\right)^2\OmK^2\notag,  
\end{align}
where the second equality assumes $c_s=c_s(r)$ and the final equality
applies to the equilibria in \S\ref{eqm}. Thus,  
\begin{align*}
N_z\lesssim
O\left(\sqrt{\epsilon_0}\OmK\right). 
\end{align*}
However, for well-mixed dust layers such that $H_\epsilon \gg \Hg$, 
$\mathrm{max}\left(N_z\right)$ may occur outside a finite vertical domain.

@@

In this section we discuss the basic properties of the 
dusty-gas mixture using Eqs.\ref{masseq}--\ref{momeq} with 
Eq. \ref{eff_energy} in place of the dust continuity
equation. Self-gravity is neglected here. Our discussion is based on
the corresponding axisymmetric linearized equations. A variable $X$ is
subject to Eulerian perturbations of the form  
\begin{align}
 \real\left[ \delta X(r,z)\exp{\left(-\ii\sigma t\right)}\right], 
\end{align}
where $\sigma$ is the complex mode frequency. For later convenience we
write 
\begin{align}
  \sigma = \ii s - \omega,
\end{align}
where $s$ and $\omega$ are the growth rate and real frequencies,
respectively. Then perturbations have time dependence $e^{st +
  \ii\omega t}$.  Thus for $\omega>0$, perturbations rotate
anti-clockwise in the complex plane.

@@

In this case, the last two integrals in Eq. \ref{int_rel} vanish. 
The condition for axisymmtric \emph{stability}, $\sigma^2>0$, 
is met if the integrand cofficients satisfy $A+D>0$ and $AD-B^2>0$
\citep[][Section 11.6]{ogilvie16}, which becomes 
\begin{align}
  \kappa^2 + c_s^2 \nabla\ln{\rhog}\cdot\nabla\tepsilon &> 0,\label{dusty_solberg1}  \\
  -c_s^2\frac{\p\ln{\rhog}}{\p z}\left(-\kappa^2\frac{\p\tepsilon}{\p
    z} + r\frac{\p\Omega^2}{\p z}\frac{\p \tepsilon}{\p r} \right) & > 0, \label{dusty_solberg2}
\end{align} 
%\begin{align}
%  \kappa^2 - \frac{1}{\rho}\nabla P \cdot \nabla s &> 0, \label{dusty_solberg1}    \\
%  -\frac{1}{\rho}\frac{\p P}{\p z} \left(\kappa^2 \frac{\p s}{\p z} -
%  r\frac{\p\Omega^2}{\p z}\frac{\p s}{\p r}\right) &>0, \label{dusty_solberg2}
%\end{align} 
where %$s = \ln{P/\rho}$, 
$\kappa^2 \equiv r^{-3}\p_r\left(r^4\Omega^2\right)>0$ is the 
square of the epicylic frequency. %We assume $\kappa^2>0$.  
Eq. \ref{dusty_solberg1}---\ref{dusty_solberg2} are in fact the 
the Solberg-Hoiland criteria for axisymmetric stability of adiabatic
gas if one inserts $C_p\ln{\left(1 -
    \tepsilon\right)}$ for the entropy into the standard
expression for said criteria \citep{tassoul78}.

We expect Eq. \ref{dusty_solberg1} to be satisfied in typical
protoplanetary disks where the dust-to-gas ratio increases in the same
direction as the local pressure gradient, which is equivalent to the
gas density gradient for isothermal gas. {\bf ref for drift} Thus for
instability we need to violate Eq. \ref{dusty_solberg2}. We explore
this below.

\subsubsection{Vertically well-mixed dust}\label{vert_mixed_theory}
In the special case where the dust-to-gas ratio is
uniform in $z$ the LHS of Eq. \ref{dusty_solberg2} becomes
\begin{align}
-\frac{c_s^4}{(1+\tepsilon)^4}\left(\frac{\p\ln{\rhog}}{\p
    z}\right)^2\left(\frac{\p\epsilon}{\p r}\right)^2<0.
\end{align}
Thus the disk is \emph{unstable}, because there is no effective vertical 
buoyancy to stabilize vertical motions ($N_z=0$). This situation is 
identical to the pure gas, adiabtic simulations of \cite{nelson13}
where the gas entropy is vertically uniform but has a radial
gradient. The authors indeed find instability. We give a numerical
example in the dusty context in \S\ref{vert_mixed}.

\subsubsection{Application to protoplanetary disks} 
%In order to apply the stability criteria
%Eq. \ref{dusty_solberg1}---\ref{dusty_solberg2} to typical
%protoplanetary disks 
We now consider the specific equilibria described in 
\S\ref{eqm}. We assume the disk is approximately Keplerian so that
$\kappa\simeq\OmK$, and that the vertical gas distribution is   
Gaussian. In terms of the dust-to-gas
ratio $\epsilon$, Eq. \ref{dusty_solberg2}
becomes 
\begin{align}
  % &1 + \frac{\epsilon}{\left(1+
  %   \epsilon\right)^2}\left(h_\mathrm{g}^2\frac{\p\ln{\rhog}}{\p\ln{r}}
  % \frac{\p\ln{\epsilon}}{\p\ln{r}} + 
  % \frac{z^2}{\Htilde^2}\right)>0 \label{stability_est1},   \\ 
&1 - \frac{\epsilon
  h_\mathrm{g}^2}{\left(1+\epsilon\right)^2}
  \frac{\p\ln{\epsilon}}{\p\ln{r}}\left(-\frac{\p\ln{\rhog}}{\p\ln{r}}+\frac{\Htilde^2}{H_\mathrm{g}^2}\frac{\p\ln{\epsilon}}{\p\ln{r}}\right)
  > 0 \label{stability_est2},
\end{align}
for \emph{stability}. The second term is usually stabilizing in 
PPDs. The third term is always destabilizing, but is small in smooth,
thin disks with radial gradients $O(1/r)$,  $h_\mathrm{g}\ll 1$,
provided that $H_\epsilon/\Hg$ is not large (e.g. some dust settling
has occured). This means that typical dusty PPDs are stable to
axisymmetric perturbations, a result valid for arbitrarily thin dust
layers.

To violate Eq. \ref{stability_est2} and obtain instability, notice the 
LHS is a quadratic in $\p_r\epsilon$. Thus instability is possible for
sufficiently large (in magnitude) radial gradients  in the dust-to-gas
ratio,  
\begin{align}
  \frac{\p\ln{\epsilon}}{\p\ln{r}} > \chi_+ \quad \mathrm{or} \quad 
  \frac{\p\ln{\epsilon}}{\p\ln{r}} < \chi_-,
\end{align}
for \emph{instability}, where
\begin{align}\label{spm}
\chi_\pm = \frac{1}{2}\frac{H_\mathrm{g}^2}{\Htilde^2} 
  \left[
  \frac{\p\ln{\rhog}}{\p\ln{r}} \pm 
  \sqrt{
  \left(\frac{\p\ln{\rhog}}{\p\ln{r}}\right)^2 + 
  4 \frac{\Htilde^2}{H_\mathrm{g}^2}
  \frac{\left(1+\epsilon\right)^2}{\epsilon h_\mathrm{g}^2}
  }
  \,\right]. 
\end{align} 
In typical accretion disks where $\p_r\rhog<0$, instability is easier
for increasing dust-to-gas ratios ($\p_r\epsilon > 0$).

We can neglect density gradients in Eq. \ref{spm} 
if $r\p_r\ln{\rhog}\sim O(1)$ and $H_\epsilon/\Hgas\gg
\sqrt{\epsilon}\hgas$. For example, if $\epsilon\simeq 0.01$ and
$\hgas\simeq 0.05$, then we require $H_\epsilon/\Hg\gg
5\times10^{-3}$. (So the dust layer thickness $\Hd\gg 5
\times10^{-3}\Hgas$ as well.) Then 
instability requires 
\begin{align}\label{dust_edge}
\left|\frac{\p\ln{\epsilon}}{\p r}\right| \gtrsim
  \frac{1}{\Htilde}\frac{\left(1+\epsilon\right)}{\sqrt{\epsilon}}.
%\simeq
 % \frac{1}{\sqrt{\epsilon}H_\epsilon}, 
\end{align}
%where the second equality applies to $\epsilon\ll 1$. 
That is, if the radial lengthscale of the dust-to-gas ratio,
$L_\epsilon\ll O(\Htilde)$, i.e. its vertical lengthscale,  
then the system is potentially unstable.  
Taking $\epsilon\sim 0.01$,  we find that 
for thin dust layers with $H_\epsilon \simeq \Hd\ll \Hg$,  instability
requires $L_\epsilon\ll O(0.1\Hg)$, i.e. the dust-to-gas ratio must 
vary on an extremely short lengthscale. This might be achieved, for
example, at sharp edges associated with gaps opened by giant planets. 
{\bf mention this is equivalent to sharp entropy gradients?}

\subsection{Thermodynamics of dust-drag instabilities}\label{dust_work}
Now consider constant $c_s$ but $\tstop\neq0$. 
Eq. \ref{int_rel} indicates that $\sigma^2$ is generally complex, so
we may have growing oscillations, or overstability, due to dust-gas
friction. This is seen by taking the imaginary part of
Eq. \ref{int_rel} 
\begin{align}
%  s = \frac{\imag\int \left(\nabla\cdot\dd\bm{v}^*\right)\dd\mathcal{C}dV}{2\omega\int\rho\left(|\dd
%    v_r|^2 + |\dd v_z|^2\right)dV}, \label{thermal_instability}
  s = \frac{\imag\int \left(\nabla\cdot\dd\bm{v}^*\right)\dd\mathcal{C}dV}{2\omega\mathcal{I}^2}, \label{thermal_instability}
\end{align} 
assuming $\omega\neq0$. %This represents overstability due to dust-gas friction. 
Since drag ($\delta \mathcal{C}$) appears as a source term in our
effective energy equation, the quantity 
$\imag\left[\left(\nabla\cdot\dd\bm{v}^*\right)\dd\mathcal{C}\right]$
may be thought of as a correlation between compression/expansion and  
heating/cooling.

@@

\subsubsection{Physical property of dust-gas drag overstabilities}

@@

A parcel of the isothermal dusty-gas mixture does 
postive work if  
\begin{align*}
-\sgn\left(\omega\right)\imag\left(\Delta\rhog\Delta\rhod^*\right)>0,
\end{align*}
meaning that gas follows dust (Fig. \ref{lag_cartoon}). Instabilities
are thus not possible if the gas does not respond to dust-drag
(i.e. no back-reaction). The pressure-density lag shown in
Fig. \ref{lag_cartoon} is achieved if, after a parcel reaches maximum
total density, its  gas content is still increasing (see $A$ to $B$ in
Fig. \ref{pdv_cartoon}). This in turn implies a sufficiently large
particle flux \emph{out} of the parcel.

@@

This thermodynamic interpretation does not explain 
\emph{why} drag forces causes gas pressure to lag behind the dust
density, but shows 
that this \emph{must} be the case for any growing oscillations associated the
dust-gas drag. To rigorously understand how dust-drag causes this 
lag requires an explicit solution to the linearized equations with 
detailed treatment of the function $\mathcal{C}$. However, given the complexity of
$\mathcal{C}$ (see Appendix \ref{lin_dust}), we might generally expect
 dusty disks to  support a range of stable and overstable modes, with the
latter being associated with pressure-density lag. 
%Indeed, \cite{jacquet11} explains the essence of the streaming 
%instability in dusty protoplanetary disks as dust accumulation
%(and hence compression) at a pressure bump, which then drags the 
%gas towards it (and hence heating) to strengthen the pressure
%bump. That is, heating occurs upon compression, as in stellar  
%pulsational instabilities \citep{cox67}.
 In \S\ref{si} we check that the
streaming instability fits into this thermodynamic interpretation in
the strong drag limit.
